---
title: "Yasmiiin"
author: "Yasmiiin"
date: "2025-03-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 

```{r }
# Load necessary libraries
library(phyloseq)

# Load the phyloseq object
load("physeq_wl.RData")

# 1. Check total number of samples
total_samples <- nsamples(physeq)
cat("Total number of samples:", total_samples, "\n\n")

# 2. Check category distribution for each environmental variable
# List of categorized variables from metadata
categories <- c("ph", "calcium", "total_nitrogen_percent", 
                "total_carbon_percent", "respiration")

# Create frequency tables for each category
invisible(
lapply(categories, function(var) {
  cat("Category distribution for", var, ":\n")
  tbl <- table(sample_data(physeq)[[var]], useNA = "ifany")
  print(tbl)
  cat("\n")
})
)
```



## We see an extreme inbalnace in Calcium so we will remove this category from our analysis

```{r}
# Remove calcium from variables to analyze
categories <- categories[categories != "calcium"]
```

## Alpha Diversity 
  Given the remaining variables (pH, nitrogen, carbon, respiration)

```{r alpha-diversity, warning=FALSE, message=FALSE}
# Load required packages
library(phyloseq)
library(vegan)
library(ggplot2)
library(ggpubr)
library(dplyr)
library(tidyr)
library(gridExtra)

# Load phyloseq object
load("physeq_wl.RData")

# Load tree file (Newick format)
tree <- read_tree("tree.nwk")

# Add to phyloseq object
physeq <- merge_phyloseq(physeq, tree)

# 1. Calculate alpha diversity metrics
alpha_metrics <- estimate_richness(physeq, 
                                  measures = c("Observed", "Shannon", "Chao1"))

# 2. Merge with metadata
meta <- data.frame(sample_data(physeq))
alpha_merged <- cbind(alpha_metrics, meta)

# 3. Define variables to analyze (only pH and nitrogen)
variables <- c("ph", "total_nitrogen_percent")

# 4. Statistical testing function
run_alpha_tests <- function(data, variables) {
  results <- list()
  
  for(var in variables) {
    n_categories <- length(unique(data[[var]]))
    
    test_func <- if(n_categories > 2) kruskal.test else wilcox.test
    test_name <- if(n_categories > 2) "Kruskal-Wallis" else "Wilcoxon"
    
    tests <- list(
      Observed = test_func(Observed ~ get(var), data = data),
      Shannon = test_func(Shannon ~ get(var), data = data),
      Chao1 = test_func(Chao1 ~ get(var), data = data)
    )
    
    results[[var]] <- list(test_name = test_name, results = tests)
  }
  return(results)
}

# Run statistical tests
test_results <- run_alpha_tests(alpha_merged, variables)

# 5. Visualization function
create_alpha_plots <- function(data, variables) {
  plot_list <- list()
  
  long_data <- data %>%
    pivot_longer(
      cols = c(Observed, Shannon, Chao1),
      names_to = "metric",
      values_to = "value"
    )
  
  metric_labels <- c(
    Observed = "Observed Species",
    Shannon = "Shannon Diversity",
    Chao1 = "Chao1 Index"
  )
  
  for(var in variables) {
    n_groups <- length(unique(data[[var]]))
    
    p <- ggplot(long_data, aes(x = .data[[var]], y = value)) +
      geom_boxplot(aes(fill = .data[[var]]), show.legend = FALSE) +
      facet_wrap(
        ~ metric, 
        scales = "free_y",
        nrow = 1,
        labeller = labeller(metric = metric_labels)
      ) +
      stat_compare_means(
        method = ifelse(n_groups > 2, "kruskal", "wilcox"),
        label = "p.format",
        label.y.npc = "top",
        size = 3.5
      ) +
      labs(
        title = paste("Alpha Diversity by", toupper(var)),
        x = gsub("_", " ", var),
        y = "Diversity Value"
      ) +
      theme_bw(base_size = 12) +
      theme(
        axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(hjust = 0.5, face = "bold"),
        strip.text = element_text(size = 11),
        panel.spacing = unit(1.5, "lines")
      )
    
    plot_list[[var]] <- p
  }
  return(plot_list)
}

# Generate plots
alpha_plots <- create_alpha_plots(alpha_merged, variables)

# Arrange final plot grid
final_plot <- ggarrange(
  plotlist = alpha_plots,
  ncol = 1,
  nrow = length(variables),
  labels = paste0(LETTERS[1:length(variables)], ")"),
  heights = rep(1.2, length(variables))
)

# Add global title and save
final_plot <- annotate_figure(final_plot,
  top = text_grob("Alpha Diversity by pH and Nitrogen Levels",
                 color = "black",
                 face = "bold",
                 size = 18))

ggsave("alpha_diversity_results.png",
       plot = final_plot,
       width = 14,
       height = 8,
       dpi = 300)

# Print statistical results
cat("\n===== ALPHA DIVERSITY STATISTICAL RESULTS =====\n")
for(var in names(test_results)) {
  cat("\n====", toupper(var), "====\n")
  cat("Test Used:", test_results[[var]]$test_name, "\n")
  print(test_results[[var]]$results)
}
```

## we need to combine the tree with the physeq object

```{r}
# Load tree file (Newick format)
tree <- read_tree("tree.nwk")

# Add to phyloseq object
physeq <- merge_phyloseq(physeq, tree)
```

# Beta Diversity Analysis

## Load Packages and Data
```{r load-data}
library(phyloseq)
library(vegan)
library(ggplot2)
library(ggpubr)
library(ape)


# Verify tree exists
if(is.null(phy_tree(physeq, errorIfNULL = FALSE))) {
  stop("Phylogenetic tree not found in phyloseq object!")
}
```

## Compute Distance Matrices
```{r compute-distances}
# Calculate phylogenetic distances
w_unifrac <- distance(physeq, method = "wunifrac")  # Weighted UniFrac
uw_unifrac <- distance(physeq, method = "unifrac")   # Unweighted UniFrac
```

## Ordination Analysis (PCoA)
```{r perform-ordination}
# Principal Coordinates Analysis
pcoa_w <- ordinate(physeq, method = "PCoA", distance = w_unifrac)
pcoa_uw <- ordinate(physeq, method = "PCoA", distance = uw_unifrac)
```

## Visualization Function
```{r ordination-plot-function}
generate_ordination <- function(ord_obj, dist_mat, title, var_group) {
  
  # Create base plot
  p <- plot_ordination(physeq, ord_obj, color = var_group) +
    geom_point(size = 3, alpha = 0.8) +
    stat_ellipse(level = 0.8, linetype = 2, linewidth = 0.5) +
    scale_color_brewer(palette = "Set1") +
    labs(
      title = paste(title, "-", var_group),
      caption = paste0("PCoA Axis 1 (", round(100*ord_obj$values$Relative_eig[1], 1), "%), ",
                      "Axis 2 (", round(100*ord_obj$values$Relative_eig[2], 1), "%)")
    ) +
    theme_bw(base_size = 12) +
    theme(
      plot.title = element_text(hjust = 0.5, face = "bold"),
      legend.position = "right"
    )
  
  # Add PERMANOVA results
  metadata <- data.frame(sample_data(physeq))
  permanova <- adonis2(dist_mat ~ metadata[[var_group]], permutations = 999)
  
  p + annotate("text", x = -Inf, y = Inf,
               label = paste("PERMANOVA\np =", signif(permanova$`Pr(>F)`[1], 3),
                            "R² =", round(permanova$R2[1], 3)),
               hjust = -0.1, vjust = 1.5, size = 3.5)
}
```

## Generate and Save Plots
```{r create-plots}
# Create plots for nitrogen and pH
ph_plot_w <- generate_ordination(pcoa_w, w_unifrac, "Weighted UniFrac", "ph")
ph_plot_uw <- generate_ordination(pcoa_uw, uw_unifrac, "Unweighted UniFrac", "ph")

nitrogen_plot_w <- generate_ordination(pcoa_w, w_unifrac, "Weighted UniFrac", "total_nitrogen_percent")
nitrogen_plot_uw <- generate_ordination(pcoa_uw, uw_unifrac, "Unweighted UniFrac", "total_nitrogen_percent")

# Arrange plots in grid
final_grid <- ggarrange(
  ph_plot_w, ph_plot_uw,
  nitrogen_plot_w, nitrogen_plot_uw,
  ncol = 2, nrow = 2,
  labels = c("A", "B", "C", "D"),
  font.label = list(size = 14)
)

# Add global title
final_grid <- annotate_figure(final_grid,
  top = text_grob("Beta Diversity by pH and Nitrogen Levels",
                 face = "bold", size = 16))

# Save high-resolution plot
ggsave("beta_diversity_plots.png", final_grid,
       width = 12, height = 10, dpi = 300,
       bg = "white")
```

## PERMANOVA Statistical Testing
```{r permanova-tests}
# Focus on nitrogen and pH only
variables_to_test <- c("ph", "total_nitrogen_percent")

# Function to test selected variables
run_permanova <- function(dist_mat) {
  results <- list()
  metadata <- data.frame(sample_data(physeq))
  
  for(var in variables_to_test) {
    formula <- as.formula(paste("dist_mat ~", var))
    results[[var]] <- adonis2(formula, data = metadata, permutations = 999)
  }
  return(results)
}

# Run tests for both distance metrics
permanova_results <- list(
  Weighted_UniFrac = run_permanova(w_unifrac),
  Unweighted_UniFrac = run_permanova(uw_unifrac)
)

# Print formatted results
cat("## PERMANOVA Results\n")
for(method in names(permanova_results)) {
  cat("\n###", method, "\n")
  print(permanova_results[[method]])
}
```

## Interpretation Notes

1. **Plot Organization**:
   - A/B: pH effects
   - C/D: Nitrogen level effects
   - Left column: Weighted UniFrac
   - Right column: Unweighted UniFrac

2. **File Output**:
   - Saved as `beta_diversity_plots.png` in working directory
   - 300 DPI resolution suitable for publication
   - 12x10 inch dimensions for clear visualization

3. **Key Comparisons**:
   - Compare weighted vs unweighted results
   - Note which variable shows stronger effects (higher R² values)
   - Check consistency between pH and nitrogen patterns
```

## Beta Diversity Statistical Analysis

```{r }
## Load Required Packages
library(phyloseq)
library(vegan)
library(ggplot2)
library(ggpubr)

## Compute Distance Matrices
w_unifrac <- distance(physeq, method = "wunifrac")
uw_unifrac <- distance(physeq, method = "unifrac")

## Statistical Testing Function
run_beta_tests <- function(dist_mat, method_name) {
  
  results <- list()
  metadata <- data.frame(sample_data(physeq))
  
  # Variables to test
  variables <- c("ph", "total_nitrogen_percent")
  
  for(var in variables) {
    # PERMANOVA
    permanova <- adonis2(dist_mat ~ metadata[[var]], permutations = 999)
    
    # Homogeneity of dispersion (Betadisper)
    dispersion <- betadisper(dist_mat, group = metadata[[var]])
    anova_disp <- anova(dispersion)
    
    # Store results
    results[[var]] <- list(
      permanova = permanova,
      dispersion = list(
        anova = anova_disp,
        distances = dispersion$distances
      )
    )
  }
  
  return(list(method = method_name, results = results))
}

## Run Tests for Both Distance Metrics
beta_results <- list(
  run_beta_tests(w_unifrac, "Weighted UniFrac"),
  run_beta_tests(uw_unifrac, "Unweighted UniFrac")
)

## Visualization: Dispersion Distances
generate_dispersion_plots <- function(disp_data, var_name) {
  ggplot(disp_data, aes(x = group, y = distances, fill = group)) +
    geom_boxplot(show.legend = FALSE) +
    labs(title = paste("Dispersion Distances -", var_name),
         x = var_name, y = "Distance to Centroid") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}

## Create and Save Dispersion Plots
disp_plots <- list()
for(method in beta_results) {
  for(var in names(method$results)) {
    plot_data <- data.frame(
      group = names(method$results[[var]]$dispersion$distances),
      distances = method$results[[var]]$dispersion$distances
    )
    
    disp_plots[[paste(method$method, var)]] <- 
      generate_dispersion_plots(plot_data, var)
  }
}

# Arrange and save dispersion plots
disp_grid <- ggarrange(plotlist = disp_plots, ncol = 2, nrow = 2)
ggsave("dispersion_plots.png", disp_grid, 
       width = 12, height = 10, dpi = 300)

## Print Statistical Results
cat("\n=== BETA DIVERSITY STATISTICAL RESULTS ===\n")
for(method in beta_results) {
  cat("\n## Method:", method$method, "\n")
  
  for(var in names(method$results)) {
    cat("\n### Variable:", var, "\n")
    
    # PERMANOVA results
    cat("- PERMANOVA:\n")
    cat("  R² =", round(method$results[[var]]$permanova$R2[1], 3), "\n")
    cat("  p-value =", method$results[[var]]$permanova$`Pr(>F)`[1], "\n")
    
    # Dispersion results
    cat("- Homogeneity of Dispersion:\n")
    cat("  F-value =", round(method$results[[var]]$dispersion$anova$`F value`[1], 2), "\n")
    cat("  p-value =", method$results[[var]]$dispersion$anova$`Pr(>F)`[1], "\n")
  }
}
```
## Interpretation Notes

1. **PERMANOVA**:
   - Significant p-value (<0.05) = communities differ between groups
   - R² value shows effect size (proportion of variance explained)

2. **Dispersion Test**:
   - Significant p-value = unequal spread between groups
   - Non-significant = valid PERMANOVA centroid comparison

3. **Key Considerations**:
   - Combine PERMANOVA and dispersion results
   - Significant PERMANOVA + non-sig dispersion = true compositional differences
   - Significant PERMANOVA + sig dispersion = differences in spread and/or composition
```

# Beta Diversity Statistical Interpretation

## Key Findings

### 1. Weighted UniFrac Results:
```{r, echo=FALSE}
# pH Analysis
cat("**pH Groups**\n")
cat("- Significant community differences (p = 0.001)\n")
cat("- Small effect size (R² = 6%)\n")
cat("- Equal group dispersion (p = 0.88)\n\n")

# Nitrogen Analysis
cat("**Nitrogen Groups**\n")
cat("- Significant community differences (p = 0.001)\n")
cat("- Small effect size (R² = 5.8%)\n") 
cat("- Equal group dispersion (p = 0.38)")
```

### 2. Unweighted UniFrac Results:
```{r, echo=FALSE}
# pH Analysis
cat("**pH Groups**\n")
cat("- Significant community differences (p = 0.001)\n")
cat("- Very small effect size (R² = 3.5%)\n")
cat("- *Unequal dispersion detected (p = 0.004)*\n\n")

# Nitrogen Analysis
cat("**Nitrogen Groups**\n")
cat("- Significant community differences (p = 0.001)\n")
cat("- Very small effect size (R² = 3.2%)\n")
cat("- Equal group dispersion (p = 0.45)")
```

## Overall Interpretation

1. **Community Composition Differences**:
   - Both pH and nitrogen show statistically significant differences in microbial communities (all p < 0.001)
   - Weighted UniFrac explains slightly more variance than Unweighted

2. **Dispersion Patterns**:
   - **pH** shows unequal dispersion in Unweighted UniFrac (p = 0.004), suggesting presence/absence patterns vary more between pH groups
   - **Nitrogen** groups have similar dispersion in both metrics (p > 0.05)

3. **Effect Size Context**:
   - All R² values < 10% indicate environmental variables explain small but significant portions of variance
   - Weighted UniFrac (abundance-weighted) shows stronger associations than Unweighted

## Recommendations

- Consider pH's dispersion effect when interpreting Unweighted UniFrac results
- Pair with alpha diversity results to understand full ecological picture
- Explore additional variables that might explain more community variance

```{r visualization, warning=FALSE, message=FALSE, fig.width=12, fig.height=8}
# Visualization of Significant Results

## 1. Alpha Diversity Boxplots
# --------------------------------------------------
# Create boxplot function
alpha_boxplot <- function(data, var, title) {
  data %>%
    pivot_longer(cols = c(Observed, Shannon, Chao1)) %>%
    ggplot(aes(x = .data[[var]], y = value)) +
    geom_boxplot(aes(fill = .data[[var]])) +
    facet_wrap(~name, scales = "free_y", 
               labeller = as_labeller(c(
                 Observed = "Observed Species",
                 Shannon = "Shannon Diversity",
                 Chao1 = "Chao1 Index"))) +
    stat_compare_means(method = "kruskal.test", label = "p.format") +
    labs(title = paste("Alpha Diversity by", toupper(title)),
         x = "", y = "Diversity Value") +
    theme_bw() +
    theme(legend.position = "none",
          axis.text.x = element_text(angle = 45, hjust = 1))
}

# Generate and save plots
ph_alpha <- alpha_boxplot(alpha_merged, "ph", "pH")
nitrogen_alpha <- alpha_boxplot(alpha_merged, "total_nitrogen_percent", "Nitrogen")

ggarrange(ph_alpha, nitrogen_alpha, ncol = 1) %>%
  ggsave("alpha_diversity_boxplots.png", ., width = 12, height = 10, dpi = 300)

## 2. Beta Diversity PCoA Plots (Corrected)
# --------------------------------------------------
# Modified PCoA plotting function
pcoa_plot <- function(ord, var, method_name) {
  plot_ordination(physeq, ord, color = var) +
    geom_point(size = 3) + 
    stat_ellipse(level = 0.8) +
    scale_color_brewer(palette = "Set1") +
    labs(title = paste(method_name, "-", var),
         subtitle = paste("PERMANOVA p =", 
                         beta_results[[which(sapply(beta_results, function(x) x$method == method_name))]]$results[[var]]$permanova$`Pr(>F)`[1])) +
    theme_bw()
}

# Generate PCoA plots
p1 <- pcoa_plot(pcoa_w, "ph", "Weighted UniFrac")
p2 <- pcoa_plot(pcoa_uw, "ph", "Unweighted UniFrac")
p3 <- pcoa_plot(pcoa_w, "total_nitrogen_percent", "Weighted UniFrac")
p4 <- pcoa_plot(pcoa_uw, "total_nitrogen_percent", "Unweighted UniFrac")

ggarrange(p1, p2, p3, p4, ncol = 2, nrow = 2) %>%
  ggsave("beta_diversity_pcoa.png", ., width = 14, height = 12, dpi = 300)

## 3. Beta Diversity Distance Boxplots (Corrected)
# --------------------------------------------------
# Modified distance comparison function
beta_distance_plot <- function(dist_mat, var, title) {
  metadata <- data.frame(sample_data(physeq))
  dist_matrix <- as.matrix(dist_mat)
  
  # Create proper distance data frame
  dist_df <- data.frame(
    Distance = as.vector(dist_matrix[upper.tri(dist_matrix)]),  # Get unique pairs
    Group = metadata[[var]][as.numeric(col(dist_matrix)[upper.tri(dist_matrix)])]
  )
  
  ggplot(dist_df, aes(x = Group, y = Distance, fill = Group)) +
    geom_boxplot(show.legend = FALSE) +
    stat_compare_means(method = "kruskal.test", label = "p.format") +
    labs(title = paste(title, "Distances by", var),
         x = var, y = "Pairwise Distance") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}

# Generate and save plots
w_dist_ph <- beta_distance_plot(w_unifrac, "ph", "Weighted UniFrac")
uw_dist_ph <- beta_distance_plot(uw_unifrac, "ph", "Unweighted UniFrac")
w_dist_nitrogen <- beta_distance_plot(w_unifrac, "total_nitrogen_percent", "Weighted UniFrac")
uw_dist_nitrogen <- beta_distance_plot(uw_unifrac, "total_nitrogen_percent", "Unweighted UniFrac")

# Arrange plots
distance_grid <- ggarrange(w_dist_ph, uw_dist_ph, 
                          w_dist_nitrogen, uw_dist_nitrogen,
                          ncol = 2, nrow = 2)

# Save plot
ggsave("beta_distance_boxplots.png", distance_grid,
       width = 14, height = 12, dpi = 300)
```

# Statistical Test Selection Based on Categories

```{r stat-test-selection, warning=FALSE, message=FALSE}
library(phyloseq)
library(ggpubr)

# Load necessary data
#load("physeq_wl.RData")
metadata <- data.frame(sample_data(physeq))

# Function to select and perform appropriate test
run_category_tests <- function(data, variables) {
  results <- list()
  
  for(var in variables) {
    # Determine number of categories
    n_cats <- length(unique(data[[var]]))
    
    # Select test
    if(n_cats < 2) {
      warning(paste(var, "has less than 2 categories - skipping"))
      next
    }
    
    test_result <- if(n_cats == 2) {
      list(
        test = "Wilcoxon rank-sum",
        result = wilcox.test(Observed ~ get(var), data = data)
      )
    } else {
      list(
        test = "Kruskal-Wallis",
        result = kruskal.test(Observed ~ get(var), data = data)
      )
    }
    
    results[[var]] <- list(
      variable = var,
      n_categories = n_cats,
      test_used = test_result$test,
      p_value = test_result$result$p.value
    )
  }
  
  return(results)
}

# Example usage with alpha diversity data
alpha_merged <- cbind(estimate_richness(physeq), metadata)
variables_to_test <- c("ph", "total_nitrogen_percent")

test_results <- run_category_tests(alpha_merged, variables_to_test)

# Format results as dataframe
results_df <- do.call(rbind, lapply(test_results, data.frame))
rownames(results_df) <- NULL

# Print formatted results
knitr::kable(results_df, caption = "Statistical Test Results")
```

## Interpretation Guide

1. **Test Selection Logic**:
   - Variables with exactly 2 categories → Wilcoxon rank-sum test
   - Variables with ≥3 categories → Kruskal-Wallis test
   - Variables with <2 categories → Skipped

2. **Result Columns**:
   - `variable`: Environmental factor tested
   - `n_categories`: Number of groups/categories
   - `test_used`: Statistical test applied
   - `p_value`: Resulting p-value

3. **Visualization Integration**:
```{r test-visualization, echo=FALSE}
# Boxplot with automatic test annotation
var <- "ph"  # Example variable
n_cats <- length(unique(alpha_merged[[var]]))

ggplot(alpha_merged, aes(x = .data[[var]], y = Observed)) +
  geom_boxplot(aes(fill = .data[[var]])) +
  {if(n_cats > 2) 
    stat_compare_means(method = "kruskal.test") else 
    stat_compare_means(method = "wilcox.test")} +
  labs(title = paste("Observed Species by", toupper(var)),
       x = var, y = "Observed Species Count") +
  theme_bw()
```



