---
title: "YNMachineLearning"
author: "Yasmiiin"
date: "2025-03-25"
output: html_document
---

```{r}
# Load required packages
library(phyloseq)
library(dplyr)
library(randomForest)
library(caret)

# Load phyloseq object
load("physeq_wl.RData")   
physeq  # check object loaded correctly

# Extract OTU table as a dataframe (samples as rows)
otu_df <- as.data.frame(t(otu_table(physeq)))

# Extract sample metadata explicitly (CRITICAL STEP!)
metadata <- as.data.frame(sample_data(physeq))

# Load ISA and overlap ASVs
isa_df <- read.csv("Core_micb_indx_taxa/ISA_nitrogen_IW.csv")
low_core <- read.csv("Core_micb_indx_taxa/overlap_tax_low.csv")
high_core <- read.csv("Core_micb_indx_taxa/overlap_tax_high.csv")

core_indicator_ids <- union(low_core$ASV, high_core$ASV)

filtered_isa <- isa_df %>%
  filter(
    ASV %in% core_indicator_ids, 
    stat >= 0.8, 
    (s.low == 1 & s.high == 0) | (s.low == 0 & s.high == 1)
  )

# Selected ASVs
selected_asvs <- filtered_isa$ASV
print(selected_asvs)

# Subset OTU table to selected ASVs
otu_select <- otu_df[, selected_asvs]


# Clearly define High and Low nitrogen categories:
# Use explicit categories based on original metadata values
metadata$Nitrogen_Class <- ifelse(metadata$total_nitrogen_percent == "high", "High",
                           ifelse(metadata$total_nitrogen_percent == "low", "Low", NA))


# Keep only samples with clear nitrogen class
valid_idx <- !is.na(metadata$Nitrogen_Class)
metadata_filt <- metadata[valid_idx, ]
otu_select_filt <- otu_df[valid_idx, selected_asvs]

# Final model data frame (features + target)
model_data <- cbind(Nitrogen_Class = metadata_filt$Nitrogen_Class, otu_select_filt)

# Verify resulting table
table(model_data$Nitrogen_Class)
head(model_data)
```
```{r}
# Explicitly set Nitrogen_Class as factor (CRITICAL FIX!)
model_data$Nitrogen_Class <- factor(model_data$Nitrogen_Class, levels = c("Low", "High"))

set.seed(123)  # reproducibility
train_idx <- createDataPartition(model_data$Nitrogen_Class, p = 0.7, list = FALSE)

train_data <- model_data[train_idx, ]
test_data  <- model_data[-train_idx, ]

# Fix column names to valid R identifiers
colnames(train_data) <- make.names(colnames(train_data))
colnames(test_data) <- make.names(colnames(test_data))




# Verify class distribution
cat("Train class distribution:\n")
print(prop.table(table(train_data$Nitrogen_Class)))

cat("Test class distribution:\n")
print(prop.table(table(test_data$Nitrogen_Class)))
```

# 2. Random Forest Model Training (500 trees)

```{r}
set.seed(123)  # reproducibility
rf_model <- randomForest(Nitrogen_Class ~ ., 
                         data = train_data, 
                         ntree = 500, 
                         importance = TRUE, 
                         proximity = TRUE)

print(rf_model)
```
# 3. Model Evaluation (Accuracy and Confusion Matrix)
Evaluate the model using both training and testing datasets.

```{r}
# Predictions on training set
train_pred <- predict(rf_model, train_data)
train_cm <- confusionMatrix(train_pred, train_data$Nitrogen_Class)
cat("Training Accuracy:", train_cm$overall["Accuracy"], "\n")
print(train_cm$table)

# Predictions on test set
test_pred <- predict(rf_model, test_data)
test_cm <- confusionMatrix(test_pred, test_data$Nitrogen_Class)
cat("Test Accuracy:", test_cm$overall["Accuracy"], "\n")
print(test_cm$table)
```

# 4. Visualization of Model Results
Variable Importance plot:
The plot identifies which ASVs are most influential in predicting nitrogen class.

```{r}
# Adjust margins to clearly display long ASV IDs
par(mar=c(5, 12, 4, 2))  

# Plot variable importance horizontally (better readability)
varImpPlot(rf_model, 
           main = "Variable Importance Plot (Random Forest)",
           type = 1,            # 1 = Mean Decrease Accuracy only (cleaner)
           cex = 0.8,           # Smaller text size
           horiz = TRUE,        # Horizontal bar plot
           las = 1)             # Horizontal labels

# Reset to default margins afterward
par(mar=c(5,4,4,2))

```
```{r}
MDSplot(rf_model, train_data$Nitrogen_Class, 
        palette=c("blue", "red"), pch=19, 
        main="MDS Plot of RF Proximities (Training Data)")
```


```{r}
# Basic RF plot for error rates and class separability visualization
plot(rf_model, main="Random Forest Model Error Rate by Number of Trees")
legend("topright", legend = colnames(rf_model$err.rate), 
       col=1:ncol(rf_model$err.rate), lty=1:2, cex=0.8)
```

### Note on Class Imbalance

The dataset is imbalanced, containing more "Low" nitrogen samples (138 samples) than "High" nitrogen samples (54 samples). This imbalance might slightly bias the model toward predicting the "Low" nitrogen class. Future analyses may consider techniques such as oversampling or SMOTE to balance the dataset for improved predictive power.


### (2) Quick modeling using the entire ASV set to explicitly compare predictive power:

```{r}
# Choose the top 200 most abundant ASVs for practical full-set comparison
asv_sums <- colSums(otu_df[valid_idx, ])
top_asvs <- names(sort(asv_sums, decreasing = TRUE)[1:200])

# Prepare dataset with selected top ASVs
full_otu_data_small <- cbind(Nitrogen_Class = metadata_filt$Nitrogen_Class, 
                             otu_df[valid_idx, top_asvs])

# Ensure factor again explicitly
full_otu_data_small$Nitrogen_Class <- factor(full_otu_data_small$Nitrogen_Class, levels = c("Low","High"))

# Fix non-standard column names
colnames(full_otu_data_small) <- make.names(colnames(full_otu_data_small))

# Train/Test split (70%/30%)
set.seed(123)
small_train_idx <- createDataPartition(full_otu_data_small$Nitrogen_Class, p=0.7, list=FALSE)
small_train <- full_otu_data_small[small_train_idx, ]
small_test <- full_otu_data_small[-small_train_idx, ]

# Train RF with the smaller practical set (top 200 ASVs)
rf_small_full <- randomForest(Nitrogen_Class ~ ., data=small_train, ntree=500)

# Evaluate accuracy explicitly
small_test_pred <- predict(rf_small_full, small_test)
small_cm <- confusionMatrix(small_test_pred, small_test$Nitrogen_Class)
cat("Accuracy (Top 200 ASVs):", small_cm$overall["Accuracy"], "\n")

```
---
title: "Predictive Analysis of Nitrogen Levels Using Microbial ASVs"
author: ""
date: "2025-03-27"
output: html_document
---

# Introduction

This report outlines the procedure used to assess the predictive power of specific microbial Amplicon Sequence Variants (ASVs) for classifying wetland soil nitrogen availability (High vs Low) using a Random Forest machine learning model. ASVs were selected based on core microbiome and indicator species analysis (ISA).

# Step-by-Step Workflow

## 1. Data and Libraries

### Libraries used:

- `phyloseq`
- `dplyr`
- `randomForest`
- `caret`

### Data:

- Phyloseq object (`physeq_wl.RData`)
- ISA results (`ISA_nitrogen_IW.csv`)
- Overlapping core-indicator ASV lists:
  - Low nitrogen: `overlap_tax_low.csv`
  - High nitrogen: `overlap_tax_high.csv`

## 2. Data Extraction and Pre-processing

The Phyloseq object was loaded, and its OTU table and metadata were extracted explicitly:

- OTU table transposed to have samples as rows, ASVs as columns.
- Sample metadata extracted clearly and explicitly.

## 3. Selection of Indicator ASVs

ASVs that overlapped between the core microbiome analysis (70% prevalence, detection = 0.001) and ISA were explicitly selected:

- Combined low/high ASV lists from `overlap_tax_low.csv` and `overlap_tax_high.csv`.
- Filtered ISA results to retain ASVs with:
  - Indicator value (`stat`) ≥ 0.8.
  - Exclusive association to one nitrogen class (low or high).

### Final ASVs selected:

- c59768a0cba53f0e1141c9ba541b6cbc
- 8d20b5539f7ea093b96ad0bf482a2d34
- 0a41bf6851bcce501e9fb0c8806e876d
- 7b0d6970e21ed778b162be105e626557
- 4ed91d931cafdca8362e8bfe1e90dab3
- 6b6af0c4ef7dc65b8c044a370146ada0
- 11334291947eca314219c89eb5ab301c
- 1d9fd315edccf1f2541d8b0a589ff1e9

## 4. Dataset Preparation for Modeling

- OTU table explicitly filtered to selected ASVs.
- Nitrogen classes (High/Low) explicitly extracted from the metadata.
- Final dataset clearly labeled, yielding 54 High and 138 Low samples.

## 5. Model Training and Evaluation

### Data Splitting:

- Data split explicitly into training (70%) and test (30%) subsets, ensuring representative proportions for both nitrogen classes.

### Random Forest Model:

- Model explicitly trained on training data (500 trees).
- Nitrogen class set as a factor explicitly to ensure classification modeling.

### Model Performance:

- **Training Accuracy** explicitly computed and reported.
- **Test Accuracy** explicitly computed and reported.

## 6. Visualization of Model Performance

### Variable Importance Plot

- Explicitly visualized to show predictive strength of individual ASVs.

### Multidimensional Scaling (MDS) Plot

- Explicit visualization of class separation using proximity measures from the Random Forest model.

### Error Rate Plot

- Visualization of error rates across number of trees, explicitly showing model stability and performance.

## 7. Additional Analysis

### Taxonomic Assignment

- Selected ASVs explicitly mapped to their taxonomic identities:
  - Domain
  - Phylum
  - Class
  - Order
  - Family
  - Genus

### Class Imbalance Noted

The dataset explicitly showed imbalance (138 Low vs 54 High samples). Future analysis may consider balancing techniques.

### Full-ASV Set Modeling Comparison

Due to memory constraints, explicit modeling with all ASVs was impractical. Therefore, the top 200 most abundant ASVs were explicitly selected for comparative modeling:

- A separate Random Forest model explicitly trained on these 200 ASVs.
- Accuracy explicitly compared with the original 8-ASV model.

## 8. Summary and Conclusions

The Random Forest modeling clearly demonstrated strong predictive capacity of the selected ASVs for classifying nitrogen availability. Explicit visualizations further supported the clear distinction between classes and highlighted the importance of selected ASVs. Documented limitations (class imbalance, computational constraints) explicitly suggest areas for potential improvement.

This structured analytical approach supports microbial community analysis as a robust method for assessing environmental nutrient availability.



